---
title: "Spatial Transcriptomics Data Analysis Tutorial 04: Clustering"
description: "This tutorial covers the preparation steps for spatial transcriptomics data analysis using Bioconductor."
author:
  - name: Youngjun Kim
    url: https://yjkim93.github.io
    orcid: 0000-0003-3346-9457
date: 12-21-2025
categories: [spatial-transcriptomics, bioconductor, R]
citation:
  url: https://yjkim93.github.io/posts/2025-12-21-spatial-transcriptomics-04-clustering/
draft: false
execute:
  warning: false
editor: source
---

# 4. Clustering and Annotation

## Load packages

```{r}
#| label: load-packages-preparation
#| message: false
library(Banksy)
library(igraph)
library(ggspavis)
library(BayesSpace)
library(nnSVG)
library(HDF5Array)
library(patchwork)
library(SpatialExperiment)
library(scran) 
library(scater)
library(scrapper)
library(SingleR)
library(qs2)
library(viridis)
library(RColorBrewer)
library(pheatmap)
```

## Load Data

```{r}
#| label: define-path
#| include: false
data_path <- "/home/kim/workspace/spatial_transcriptomics/data"
```

```{r}
#| label: define-data-path
visium_data_path <- file.path(data_path, "raw_colon_cancer")
hdf_data_path <- file.path(data_path, "hdf5_colon_cancer")
```

```{r}
#| label: import-data
spe <- loadHDF5SummarizedExperiment(
  dir = hdf_data_path,
  prefix = "03_processed_"
)
```

## Clustering

### Non-spatially aware clustering

#### Leiden Clustering

The simplest way to cluster the spots is to use the same methods as in scRNA-seq analysis. Here, we use the Leiden clustering algorithm on the PCs obtained from standard PCA. We use the `scran::buildSNNGraph()` function to build a shared nearest neighbor (SNN) graph based on the Jaccard similarity of the k-nearest neighbors in PCA space. We then apply the Leiden algorithm using `igraph::cluster_leiden()` function to identify clusters of spots.

```{r}
#| label: run-leiden-clustering

# Build shared nearest neighbor (SNN) graph
snn_graph <- buildSNNGraph(spe, use.dimred = "PCA", type="jaccard", k = 20)

# Cluster using the Leiden algorithm
leiden_cluster <- cluster_leiden(snn_graph, objective_function="modularity", resolution = 0.6)

# Add cluster labels to colData
spe$leiden_cluster <- factor(leiden_cluster$membership)

table(spe$leiden_cluster)
```

### Spatially aware clustering
For spatial transcriptomics, we can also incorporate spatial information into the clustering process. Here, we will use two methods: `BayesSpace` and `BANKSY`.

#### Clustering with BayesSpace

[`BayesSpace`](https://www.nature.com/articles/s41587-021-00935-2) uses a Bayesian statistical model to cluster spots based on both gene expression and spatial location.

```{r}
#| label: run-bayesspace-clustering

# Prepare data for 'BayesSpace'. We skip PCA here as we have already computed PCs.
spe <- spatialPreprocess(spe, skip.PCA=TRUE)

# Perform spatial clustering with 'BayesSpace' using 'd=10' PCs and targeting 'q=8' clusters
spe <- spatialCluster(spe, q=8, d=10, nrep=1e3, burn.in=100) 

# Add BayesSpace cluster labels to colData
spe$bayesspace_cluster <- factor(spe$spatial.cluster)
table(spe$bayesspace_cluster)
```

#### Clustering with BANKSY

Since we have already computed spatially-aware PCs using `BANKSY`, we can use these PCs to build an SNN graph and perform clustering using the Leiden algorithm. We will use the PCs obtained from BANKSY with λ = 0.2 or 0.8 for clustering.

```{r}
#| label: run-banksy-clustering
# Build shared nearest neighbor (SNN) graph
snn_graph_banksy_lam0.2 <- buildSNNGraph(spe, use.dimred = "PCA_banksy_lam0.2", type="jaccard", k = 20)
snn_graph_banksy_lam0.8 <- buildSNNGraph(spe, use.dimred = "PCA_banksy_lam0.8", type="jaccard", k = 20)

# Cluster using the Leiden algorithm
banksy_cluster_lam0.2 <- cluster_leiden(snn_graph_banksy_lam0.2, objective_function="modularity", resolution = 0.6)
banksy_cluster_lam0.8 <- cluster_leiden(snn_graph_banksy_lam0.8, objective_function="modularity", resolution = 0.6)

# Add cluster labels to colData
spe$banksy_cluster_lam0.2 <- factor(banksy_cluster_lam0.2$membership)
spe$banksy_cluster_lam0.8 <- factor(banksy_cluster_lam0.8$membership)

table(spe$banksy_cluster_lam0.2)
table(spe$banksy_cluster_lam0.8)
```

### Visualization of the different clustering results
```{r}
#| label: plot-clustering-results

ks <- c("leiden_cluster", "bayesspace_cluster", "banksy_cluster_lam0.2", "banksy_cluster_lam0.8") 

# Plot spatial distribution of clusters for each method
plots <- c(lapply(ks, \(.) {
    plt <- plotVisium(spe, annotate=., zoom = T, point_shape = 22, point_size = 1)
    plt
}) ,  
  plotVisium(spe, zoom = T,  spots=FALSE )) # add a plot of the tissue without spots

# combine plots
plots |>  wrap_plots(ncol=2) &
      scale_fill_brewer(palette = "Set1") &
    theme(legend.key.size=unit(0, "lines"), legend.justification="left")
```

We can appreciate that all methods capture major tissue structures. The borders of the clusters from `BANKSY` with λ = 0.8 appear to be more "smoothened out". This is expected as higher λ values give more weight to spatial information.

It may be helpful to visualize individual clusters to see the spatial patterns more clearly. Here, we see the clusters obtained from `BANKSY` with λ = 0.2.

```{r}
#| label: plot-individual-clusters
# plot selected clusters in order of frequency,

cluster_col <- "banksy_cluster_lam0.2"

# highlighting cells assigned to cluster 'k'
lapply(tail(names(sort(table(spe[[cluster_col]]))), length(levels(spe[[cluster_col]]))), \(k) {
    spe$foo <- spe[[cluster_col]] == k
    spe <- spe[, order(spe$foo)]
    plt <- plotCoords(spe, annotate="foo")
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.5
    plt + ggtitle(k)
}) |>
    wrap_plots(nrow=3) &
    scale_color_manual(values=c("white", "red")) &
    theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

## Annotation




## Save the processed data
```{r}
#| label: save-processed-data
saveHDF5SummarizedExperiment(
  spe,
  dir = hdf_data_path,
  prefix = "04_clustered_",
  replace = TRUE,
  chunkdim = NULL,
  level = NULL,
  as.sparse = NA,
  verbose = NA
)
```

:::{.callout-tip collapse="true"}
## Expand for Session Info
```{r}
#| label: session-info
sessionInfo()
```
:::